---
title: "Exploratory Analysis"
output: 
  html_document:
    toc: true
    toc_float: true
---
```{r include=FALSE}
library(flexdashboard)
library(knitr)
library(rmarkdown)
library(tidyverse)
library(plotly)
library(haven)
```


```{r nhats, echo=FALSE}
nhats = read_sas("./final_data/nhats_r5.sas7bdat")  %>%
  select(spid, hc5depresan1, hc5depresan2, r5dgender, r5d2intvrage, rl5dracehisp, md5canewlker, 
         sn5dnumsn, cg5ratememry, hc5health, wb5truestme4) %>%
  janitor::clean_names() %>%   
  filter(hc5depresan1 > 0, hc5depresan2 > 0, sn5dnumsn >= 0, r5d2intvrage > 0, md5canewlker > 0, 
         cg5ratememry > 0, hc5health > 0, wb5truestme4 > 0, rl5dracehisp < 5)

nhats %>% 
  select(hc5depresan1, hc5depresan2)  %>% 
  rowSums(na.rm = TRUE) -> nhats$phq.total

nhats = nhats %>%
  mutate(phq.total = phq.total - 2) %>% 
  mutate(phq_cat = if_else(phq.total < 3, "Low Depression Risk", "Elevated Depression Risk")) %>% 
  mutate(
    gender = recode(r5dgender, `1` = "Male", `2` = "Female"),
    gender = factor(gender),
    age_cat = recode(r5d2intvrage, `1` = "65 to 69", `2` = "70 to 74", `3` = "75 to 79", 
                 `4` = "80 to 84", `5` = "85 to 89", `6` = "90+"),
    age_cat = factor(age_cat, levels = c("65 to 69", "70 to 74", "75 to 79", "80 to 84", 
                  "85 to 89", "90+")),
    race_ethnicity = recode(rl5dracehisp, `1` = "White, non-hispanic", `2` = "Black, non-hispanic", 
                  `3` = "Other, non-hispanic", `4` = "Hispanic"),
    race_ethnicity = factor(race_ethnicity),
    cane_walker_wheelchair = recode(md5canewlker, `1` = "Yes", `2` = "No"),
    cane_walker_wheelchair = factor(cane_walker_wheelchair),
    memory = recode(cg5ratememry, `1` = "Excellent", `2` = "Very Good", `3` = "Good", 
                    `4` = "Fair", `5` = "Poor"),
    memory = factor(memory, c("Excellent", "Very Good", "Good", "Fair", "Poor")),
    overall_health = recode(hc5health, `1` = "Excellent", `2` = "Very Good", `3` = "Good", 
                    `4` = "Fair", `5` = "Poor"),
    overall_health = factor(overall_health, levels = c("Excellent", "Very Good", "Good", "Fair", 
                    "Poor")),
    likes_living_situation = recode(wb5truestme4, `1` = "Agree a lot", `2` = "Agree a little", 
                    `3` = "Agree not at all"),
    likes_living_situation = factor(likes_living_situation, c("Agree a lot", "Agree a little", 
                    "Agree not at all")),
    ) %>% 
  rename(social_network_size = sn5dnumsn)
```

### Living situation

Plotting of participant view of living situation predictor against PHQ-2, stratified by race (EM) to show the differences in PHQ-2 (depression risk scale) distribution across categories. 
```{r nhats_living_situation, echo=FALSE, warning=FALSE,message=FALSE}

living_plot = nhats %>% 
  ggplot(aes(x = likes_living_situation, y = phq.total, color = race_ethnicity)) + 
  geom_point(position = "jitter", size = 0.5) + 
  labs(title = "View of living situation and PHQ-2 score distribution",
       x = "Likes living situation", y = "PHQ-2 score") + 
  facet_grid(rows = vars(race_ethnicity)) + 
  theme(legend.position = "none")

ggplotly(living_plot)

```

### Social Network

Plotting social network (by race/gender/age_cat) and health by phq_cat:

```{r, echo=FALSE,warning=FALSE,message=FALSE}

sn_phq = nhats %>% 
  ggplot(aes(x = social_network_size, group = phq_cat)) + 
    geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat = "count") +
    geom_text(aes( label = scales::percent(round(..prop.., 3)),
                   y = ..prop.. ), stat = "count", vjust = -.3, size = 2.5) +
    labs(y = "Percent", fill = "SN") +
    facet_grid(~phq_cat) +
    scale_y_continuous(labels = scales::percent)

ggplotly(sn_phq)

health_phq = nhats %>% 
  filter(hc5health > 0) %>%
  ggplot(aes(x = overall_health, group = phq_cat)) + 
    geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat = "count") +
    geom_text(aes( label = scales::percent(round(..prop.., 2)),
                   y = ..prop.. ), stat = "count", vjust = -2, size = 3) +
    labs(y = "Percent", fill = "Overall Health") +
    facet_grid(~phq_cat) +
    scale_y_continuous(labels = scales::percent)

ggplotly(health_phq)
```

### Self rated memory

####  Visualizing any differences in depression outcome by predictor, self-rated memory. 

- Survey respondents were asked to rate their memory on a scale of 1-5, 1 being Excellent and 5 being Poor. This plot is a visualization of the proportion of respondents in this dataset by their depression outcome status - Mild/No Depression or Moderate/Severe Depression. The figure is broken down by how participants rated their memory (x-axis). 

```{r, echo=FALSE,warning=FALSE,message=FALSE}
mem_plot = nhats %>%  group_by(phq_cat, memory) %>% 
  summarise(n = n()) %>% mutate(
  proportion = (n / sum(n)))

  memory_plot = mem_plot %>% mutate(memory = factor(memory)) %>% drop_na() %>%   
    ggplot(aes(x = memory, y = proportion, group = phq_cat)) + geom_line(alpha = 0.4) +
  
    geom_point(alpha = 0.5, color = "red") + 
    facet_grid(~phq_cat) + 
    labs( title = "Self rated memory and depression outcomes",
    x = "Self rated memory", y = " Proportion of respondents") +  
    scale_y_continuous(labels = scales::percent)
  
ggplotly(memory_plot)
```

### Use of cane walker or wheelchair

####  Visualizing any differences in depression outcome by predictor, use of cane walker wheelchair. 

- Survey respondents were asked if they used a cane walker wheelchair. This plot is a visualization of the proportion of respondents in this dataset by their depression outcome status - Mild/No Depression or Moderate/Severe Depression. The figure is broken down by if participants responded Yes, or No (x-axis). 


```{r, echo=FALSE,warning=FALSE,message=FALSE}
cane_walker_wc_plot = nhats %>%  
    group_by(phq_cat, cane_walker_wheelchair) %>% 
    summarise(n = n()) %>%  
    mutate(proportion = (n / sum(n)))

cane_walker_wc_plot = 
  cane_walker_wc_plot %>%  
  ggplot(aes(x = cane_walker_wheelchair, y =proportion)) + 
  geom_col(fill= "light blue") + facet_grid(~phq_cat)+
  labs( title = "Use of cane walker wheelchair and depression outcomes",
      x = "Asked if used a cane walker wheel chair", y = " Proportion of respondents")+  
    scale_x_discrete(
      labels = c("Yes","No")) + 
    scale_y_continuous(labels = scales::percent)

  ggplotly(cane_walker_wc_plot)
```




