---
title: "Exploratory Analysis"
output: 
  html_document:
    toc: true
    toc_float: true
---
```{r include=FALSE}
library(flexdashboard)
library(knitr)
library(rmarkdown)
library(tidyverse)
library(plotly)
library(haven)
```


```{r nhats, echo=FALSE, fig.align='center'}
nhats = read_sas("./final_data/nhats_r5.sas7bdat")  %>%
  select(spid, hc5depresan1, hc5depresan2, r5dgender, r5d2intvrage, rl5dracehisp, md5canewlker, 
         sn5dnumsn, cg5ratememry, hc5health, wb5truestme4) %>%
  janitor::clean_names() %>%   
  filter(hc5depresan1 > 0, hc5depresan2 > 0, sn5dnumsn >= 0, r5d2intvrage > 0, md5canewlker > 0, 
         cg5ratememry > 0, hc5health > 0, wb5truestme4 > 0, rl5dracehisp < 5)

nhats %>% 
  select(hc5depresan1, hc5depresan2)  %>% 
  rowSums(na.rm = TRUE) -> nhats$phq.total

nhats = nhats %>%
  mutate(phq.total = phq.total - 2) %>% 
  mutate(phq_cat = if_else(phq.total < 3, "Low Depression Risk", "Elevated Depression Risk")) %>% 
  mutate(
    gender = recode(r5dgender, `1` = "Male", `2` = "Female"),
    gender = factor(gender),
    age_cat = recode(r5d2intvrage, `1` = "65 to 69", `2` = "70 to 74", `3` = "75 to 79", 
                 `4` = "80 to 84", `5` = "85 to 89", `6` = "90+"),
    age_cat = factor(age_cat, levels = c("65 to 69", "70 to 74", "75 to 79", "80 to 84", 
                  "85 to 89", "90+")),
    race_ethnicity = recode(rl5dracehisp, `1` = "White, non-hispanic", `2` = "Black, non-hispanic", 
                  `3` = "Other, non-hispanic", `4` = "Hispanic"),
    race_ethnicity = factor(race_ethnicity),
    cane_walker_wheelchair = recode(md5canewlker, `1` = "Yes", `2` = "No"),
    cane_walker_wheelchair = factor(cane_walker_wheelchair),
    memory = recode(cg5ratememry, `1` = "Excellent", `2` = "Very Good", `3` = "Good", 
                    `4` = "Fair", `5` = "Poor"),
    memory = factor(memory, c("Excellent", "Very Good", "Good", "Fair", "Poor")),
    overall_health = recode(hc5health, `1` = "Excellent", `2` = "Very Good", `3` = "Good", 
                    `4` = "Fair", `5` = "Poor"),
    overall_health = factor(overall_health, levels = c("Excellent", "Very Good", "Good", "Fair", 
                    "Poor")),
    likes_living_situation = recode(wb5truestme4, `1` = "Agree a lot", `2` = "Agree a little", 
                    `3` = "Agree not at all"),
    likes_living_situation = factor(likes_living_situation, c("Agree a lot", "Agree a little", 
                    "Agree not at all")),
    ) %>% 
  rename(social_network_size = sn5dnumsn)
```

### Living situation

Plotting of participant view of living situation predictor against PHQ-2, stratified by race (EM) to show the differences in PHQ-2 (depression risk scale) distribution across categories. Proportion of higher PHQ-2 score were greater among lower satisfaction with living situations. 

```{r nhats_living_situation, echo=FALSE, warning=FALSE,message=FALSE}
living_plot = nhats %>% 
  ggplot(aes(x = likes_living_situation, y = phq.total, color = race_ethnicity)) + 
  geom_point(position = "jitter", size = 0.5) + 
  labs(title = "View of living situation and PHQ-2 score distribution",
       x = "Likes living situation", y = "PHQ-2 score") + 
  facet_grid(rows = vars(race_ethnicity)) + 
  theme(legend.position = "none")

ggplotly(living_plot)

```

### Perceived Overall Health

#### Visualizing any differences in depression outcome by predictor, perceived overall health.

- The respondents were asked to rate their overall health on a scale from 1-5 with 1 being Excellent health and 5 being poor health. The graph below illustrates the proportion of respondents rating their overall health stratified by their depression outcome (PHQ-2) status. We can observe that respondents who scored as Mild/No depression recorded their perceived overall health within the Excellent to Good range while respondents with Moderate/Severe depression were more likely to report their overall health within the Good to Poor range. 

```{r, echo=FALSE}
nhats %>% 
  filter(hc5health > 0) %>%
  ggplot(aes(x = overall_health, group = phq_cat)) + 
    geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat = "count") +
    geom_text(aes( label = scales::percent(round(..prop.., 2)),
                   y = ..prop.. ), stat = "count", vjust = -0.5, size = 3) +
    labs(y = "Proportion of Respondents", x = "Perceived Overall Health", fill = "Overall Health") +
    theme(legend.position = "none")  + theme(axis.text.x = element_text(angle = 30, vjust = 0.7)) +
    facet_grid(~phq_cat) + scale_y_continuous(labels = scales::percent) + 
    theme(panel.background = element_rect(fill = "white", colour = "grey50")) +
    scale_fill_viridis_d()

```

### Social Network Size

#### Visualizing any differences in depression outcome by predictor, social network size - stratified by gender.

- The graph below shows the mean social network size for different age groups stratified by gender and depression risk. From the graphs we can observe that female respondents at a low risk of depression are much likely to have a greater social network size on average as compared to females with a higher risk for depression. We also observe that females, in general, have a greater social network group as compared to males, controlling for depression risk.

```{r, echo=FALSE, message=FALSE}
sn_plot = nhats %>% group_by(phq_cat, age_cat, gender) %>% 
  summarise(
    mean = mean(social_network_size)
  ) %>%
  mutate(age_cat = factor(age_cat)) 
         
sn_plot %>%
  ggplot(aes(x = age_cat, y = mean, group = gender, color = gender)) + 
  geom_line(alpha = 0.5) + facet_grid(~phq_cat) + geom_point() +
  theme(axis.text.x = element_text(angle = 30, vjust = 0.7)) +
  theme(panel.background = element_rect(fill = "white", colour = "grey50")) +
  labs(y = "Mean Social Network Size", x = "Age Groups") + scale_color_viridis_d(name = "Gender")
```

### Self rated memory

####  Visualizing any differences in depression outcome by predictor, self-rated memory. 

- Survey respondents were asked to rate their memory on a scale of 1-5, 1 being Excellent and 5 being Poor. This plot is a visualization of the proportion of respondents in this dataset by their depression outcome status - Low depression risk or elevated depression risk. The figure is broken down by how participants rated their memory. 

Initial observations: 


- A slightly lower proportion of respondents whi rated their memory as "Excellent" who had high risk of depression symptoms compared to those who had low risk of depression symptomns and scored "Excellent".
- Higher proportion of high-risk depression respondents who rated their memory as "Poor" compared to low risk depression patients who also rated their memory as "Poor". 

```{r, echo=FALSE,warning=FALSE,message=FALSE}
mem_plot = nhats %>%  group_by(phq_cat, memory) %>% 
  summarise(n = n()) %>% mutate(
  proportion = (n / sum(n)))

  memory_plot = mem_plot %>% mutate(memory = factor(memory)) %>% drop_na() %>%   
    ggplot(aes(x = memory, y = proportion, group = phq_cat)) + geom_line(alpha = 0.4) +
  
    geom_point(alpha = 0.7, color = "light blue") + 
    facet_grid(~phq_cat) + 
    labs( title = "Self rated memory and depression outcomes",
    x = "Self rated memory", y = " Proportion of respondents") +  
    theme(panel.background = element_rect(fill = "white", colour = "grey50")) +
     
    scale_y_continuous(labels = scales::percent)
  
ggplotly(memory_plot)
```

### Use of cane walker or wheelchair

####  Visualizing any differences in depression outcome by predictor, use of assistive devices (cane, walker, wheelchair). 

Survey respondents were asked if they used a cane walker wheelchair. The plot below is a visualization of the proportion of respondents in this dataset by their level of depression symptom risk. The figure is stratified by if participants responded Yes, or No to using a cane, walker, or wheelchir. 

Initial observations:


-  A higher proportion of those who responded to using an assitive device having elevated risk for depression symptoms compared to those who responded "no" using an assistive device and have low risk of depression symptoms. T
- There doesn't appear to be any differences in proportion of respondents use or don't use an assistive device among those who had elevated depression risk. Among those who had a low risk of depression scores, we see that there is a higher proportion respondents that do use an assistive device. 



```{r, echo=FALSE,warning=FALSE,message=FALSE, fig.align='center'}
cane_walker_wc_plot = nhats %>%  
    group_by(phq_cat, cane_walker_wheelchair) %>% 
    summarise(n = n()) %>%  
    mutate(proportion = (n / sum(n)))

cane_walker_wc_plot = 
  cane_walker_wc_plot %>%  
  ggplot(aes(x = cane_walker_wheelchair, y =proportion)) + 
  geom_col(fill= "light blue") + facet_grid(~phq_cat)+
  labs( title = "Use of cane walker wheelchair and depression outcomes",
      x = "Asked if used a cane, walker/wheel chair", y = " Proportion of respondents")+  
    scale_x_discrete(
      labels = c("Yes","No")) + 
    scale_y_continuous(labels = scales::percent) + 
  theme(panel.background = element_rect(fill = "white", colour = "grey50"))

  ggplotly(cane_walker_wc_plot)
```


