---
title: <center> P8105 - Final Project Report </center>
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
library(haven)
library(tidyverse)
library(plotly)


knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```


```{r, echo=FALSE}
nhats = read_sas("./final_data/nhats_r5.sas7bdat")  %>%
  select(spid, hc5depresan1, hc5depresan2, r5dgender, r5d2intvrage, rl5dracehisp, md5canewlker, 
         sn5dnumsn, cg5ratememry, hc5health, wb5truestme4) %>%
  janitor::clean_names() %>%   
  filter(hc5depresan1 > 0, hc5depresan2 > 0, sn5dnumsn >= 0, r5d2intvrage > 0, md5canewlker > 0, 
         cg5ratememry > 0, hc5health > 0, wb5truestme4 > 0, rl5dracehisp < 5)

nhats %>% 
  select(hc5depresan1, hc5depresan2)  %>% 
  rowSums(na.rm = TRUE) -> nhats$phq.total

nhats = nhats %>%
  mutate(phq.total = phq.total - 2) %>% 
  mutate(phq_cat = if_else(phq.total < 3, "Mild/No Depression", "Moderate/Severe Depression")) %>% 
  mutate(
    gender = recode(r5dgender, `1` = "Male", `2` = "Female"),
    gender = factor(gender),
    age_cat = recode(r5d2intvrage, `1` = "65 to 69", `2` = "70 to 74", `3` = "75 to 79", 
                 `4` = "80 to 84", `5` = "85 to 89", `6` = "90+"),
    age_cat = factor(age_cat, levels = c("65 to 69", "70 to 74", "75 to 79", "80 to 84", 
                  "85 to 89", "90+")),
    race_ethnicity = recode(rl5dracehisp, `1` = "White, non-hispanic", `2` = "Black, non-hispanic", 
                  `3` = "Other, non-hispanic", `4` = "Hispanic"),
    race_ethnicity = factor(race_ethnicity),
    cane_walker_wheelchair = recode(md5canewlker, `1` = "Yes", `2` = "No"),
    cane_walker_wheelchair = factor(cane_walker_wheelchair),
    memory = recode(cg5ratememry, `1` = "Excellent", `2` = "Very Good", `3` = "Good", 
                    `4` = "Fair", `5` = "Poor"),
    memory = factor(memory, c("Excellent", "Very Good", "Good", "Fair", "Poor")),
    overall_health = recode(hc5health, `1` = "Excellent", `2` = "Very Good", `3` = "Good", 
                    `4` = "Fair", `5` = "Poor"),
    overall_health = factor(overall_health, levels = c("Excellent", "Very Good", "Good", "Fair", 
                    "Poor")),
    likes_living_situation = recode(wb5truestme4, `1` = "Agree a lot", `2` = "Agree a little", 
                    `3` = "Agree not at all"),
    likes_living_situation = factor(likes_living_situation, c("Agree a lot", "Agree a little", 
                    "Agree not at all")),
    ) %>% 
    rename(social_network_size = sn5dnumsn)
```
<center>
### Assessing predictors of depression risk in older adults



##### Group Members:
Ifrah Fayyaz (if2282), Lydia Parr (lcp2148), Lauren Cohen (lcc2181), Onyeka Isamah (oi2142), Riddhi Thakkar (rt2640) 

</center> 


#### Study motivation and goals: 

The motivation for this project is to identify predictors of depression outcomes among older adults. Predictors in this project are:

* `gender`: Gender (Male, Female)
* `age_cat`: Categorical age at interview (65 to 69, 70 to 74, 75 to 79, 80-84, 85-89, 90+)
* `race_ethnicity`: Race and hispanic ethnicity (White, non-hispanic; Black, non-hispanic; Other, non-Hispanic; Hispanic)
* `cane_walker_wheelchair`: Used cane, walker, wheelchair (Yes, No)
* `memory`: Rate your memory (Excellent, Very Good, Good, Fair, Poor)
* `overall_health`: Overall health condition (Excellent, Very Good, Good, Fair, Poor)
* `likes_living_situation`: Likes living situation (Agree a lot, Agree a little, Agree not at all)
* `social_network_size`: Social network size (0, 1, 2, 3, 4, 5)

* The outcome variable of interest is:
*`PHQ-cat`: Depression (dichotomized based on total PHQ-2 score)

We chose older adults as the target population since they are a population that are known to have higher risk of depression compared to the general population. Paper's that studied the associations of these predictor variables and depression outcomes were also the inspiration of this paper. 


#### Initial questions:
We initially wanted to see if comparing various exposure groups would be signficantly associated with a respondent scoring 3 or higher on the PHQ-2 depression test for depression symptoms. 

**Data**: 


**Source**: National Health and Aging Trends Study
from Johns Hopkins and Westat

**Methods and data cleaning**: 
Using only round 5 data, included ID variable, the two variables needed for the depression risk scale (PHQ-2), gender, age, race, whether participant uses mobility devices, size of social network, whether or not they have no one to talk to, self reported memory issues causing problems, self-rated health, and whether they are happy with their living situation. 

In creating dichotomous variable phq_cat regarding depression status, phq.total was subtracted by 2 to account for the fact that total scores on the PHQ-2 should range from 0-6 (sum of two items ranging from 0-3, while scores for the two items provided in our data ranged from 1-4). Once 2 was subtracted from the total score, phq_cat was dichotomized based on a cutoff point of 3 or higher.

Categorical variables were recoded to include descriptions of values and converted into factors.


##### Exploratory analysis: 

Visualizations, summaries, and exploratory statistical analyses. Justify the steps you took, and show any major changes to your ideas.

- Figure 1: Demographic frequencies:

- Figure 2: Participant view of living situation and PHQ-2 score distribution

- Figure 3: Plotting social network  by phq_cat:

- Figure 4: Plotting health codnition by phq_cat

- Figure 5: Self rated memory and depression categories 

- Figure 6: Use of assistive devices (cane, walker, wheelchair) and depression outcomes (PHQ-2)
```{r, echo=FALSE, message= FALSE}
cane_walker_wc_plot = nhats %>%  
    group_by(phq_cat, cane_walker_wheelchair) %>% 
    summarise(n = n()) %>%  
    mutate(proportion = (n / sum(n)))

cane_walker_wc_plot = 
  cane_walker_wc_plot %>%  
  ggplot(aes(x = cane_walker_wheelchair, y =proportion)) + 
  geom_col(fill= "light blue") + facet_grid(~phq_cat)+
  labs( title = "Use of cane walker wheelchair and depression outcomes",
      x = "Asked if used a cane walker wheel chair", y = " Proportion of respondents")+  
    scale_x_discrete(
      labels = c("Yes","No")) + 
    scale_y_continuous(labels = scales::percent)

  ggplotly(cane_walker_wc_plot)
```


##### **Additional analysis**: Multivariable logisitc regression

 


##### **Discussion**: What were your findings? Are they what you expect? What insights into the data can you make?